Best Practices for Multiple Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-05-23 paper/manuscript.tex
first draft

\section{\tocorange{Large-Scale Simulation Planning}}
\label{sec:large_scale_planning}
\tocsectioncomment{Possibly add discussion of methods which are not easily characterised as single/ multi e.g. lambda dynamics? Maybe a better distinction is calculations where we are interested in more than two end states.}
\tocsectioncomment{@OpenFE team to add discussion of how to run and adaptively update perturbation networks.}
\tocsectioncomment{Possibly add discussion of different protocols, especially active learning. Best practices may not be well established but many companies are running at scale.}
\tocsectioncomment{Discussion of merits of ABFE/ RBFE for virtual screening? See @Gary Tresadern comment from first meeting - lots of discussion around ABFE calculations for virtual screening.}

\paragraph{Perturbation maps}
\tocsectioncomment{Perturbation maps paragraph: @Mary Pitman/ @Jenke Scheen/ OpenFE team to update discussion of perturbation maps?}
For RBFE campaigns with more than two ligands in the input series a \textit{perturbation map} or \textit{-network} must be generated: this is simply a set of data that describe which transformations will be executed between the ligands (referred to as \textit{edges}). \cite{MOBLEY LOMAP} Heuristics have shown the more connected the perturbation network the higher the free energy prediction accuracy versus experimental measures. Running a fully-connected network (i.e. simulate transformations between \textit{all} ligands in the campaign) is typically considered too computationally expensive and may not be suited optimally with current estimators such as MLE (\ref{SECTION MLE}).\cite{SCHEEN ET AL} There exist several methods for optimization of network structure while minimizing the number of perturbations that need to be computed, reducing the resulting computational cost at a minimal decrease in accuracy ~\cite{yang2020optimal,xu2019optimal}. Sometimes - depending on the RBFE implementation - the introduction of intermediate ligands that are not part of the original congeneric series can be helpful for avoiding challenging transformations such as ring breaking or transformations that would otherwise result in large numbers of atoms being inserted or deleted.\cite{LINDE PAPER} Some commercial tools have well-calibrated underlying heuristics to plan these network designs but may fail with complicated input datasets, thus requiring manual user validation, in particular when dealing with chiral compounds.

@@

In some cases, when dealing with for example very large datasets that would benefit from rougher initial free energy ranking or in cases where perturbations would be rather large, a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} \textbf{A} is used. However, adding redundancy into the network means that a better error analysis can be carried out by analyzing cycle closure errors as discussed in sec.~\ref{sec:are-they-good}, with an example given in Fig.~\ref{fig:fig_types_of_networks}\textbf{B}.

Methods from experimental design have been applied to the construction of perturbation maps. Yang et al.~\cite{yang2020optimal} showed how to optimize the perturbation map by selecting a fixed number of calculations from the pairwise perturbations so that the resulting set of calculations minimize the total variance. Xu~\cite{xu2019optimal} showed how to optimize the perturbation map by allocating different amounts of simulation time to different pairwise perturbations so as to minimize the total variance, given the total simulation time of all the perturbation calculations. Both approaches lead to substantial reduction in the statistical error of the estimated free energies.

>>>>

\section{\tocorange{Large-Scale Simulation Planning}}
\label{sec:large_scale_planning}
\tocsectioncomment{Possibly add discussion of methods which are not easily characterised as single/ multi e.g. lambda dynamics? Maybe a better distinction is calculations where we are interested in more than two end states.}
\tocsectioncomment{@OpenFE team to add discussion of how to run and adaptively update perturbation networks.}
\tocsectioncomment{Possibly add discussion of different protocols, especially active learning. Best practices may not be well established but many companies are running at scale.}
\tocsectioncomment{Discussion of merits of ABFE/ RBFE for virtual screening? See @Gary Tresadern comment from first meeting - lots of discussion around ABFE calculations for virtual screening.}
{\color{blue}
When tackling large datasets, methods that compute multiple ligands' binding free energies simultaneously offer an appealing alternative to pairwise calculations. These approaches can be categorized as "multiple end-state" free energy calculations (or "multistate" for short). They can in principle be more convenient than their pairwise counterparts as they sidestep ligand network construction entirely and can deliver all relative free energies in a single run. However, requiring many end states to be sampled at once complicates the energy landscape, forcing longer, inherently sequential simulations. Additionally they are often more difficult to set up: combining multiple ligand topologies raises extra challenges about avoiding bias from distance restraints or hybrid models while requiring sufficient phase-space overlap.

Enveloping distribution sampling \cite{christ2007enveloping} (EDS) works by constructing a single reference potential controlled by parameters that flatten inter-state barriers and balance end-state sampling. Multiple EDS potentials with different levels of barrier-smoothing can be combined in a replica-exchange setup in order to promote crossing the alchemical energy barriers that separate the end-states \cite{sidler2017efficient, champion2023accelerating}. This approach (RE-EDS) confers several practical advantages. Because the protein and solvent degrees of freedom are propagated only once per time step, the incremental cost of adding more ligands grows only linearly (for evaluating their potentials) rather than quadratically (in a fully connected perturbation graph). Furthermore, RE-EDS inherits intrinsic cycle closure from the enveloping reference state, eliminating the need to check or enforce thermodynamic cycles by hand.

Setting up RE-EDS simulations requires some thought to be put in system preparation. In the dual-topology setup all ligands must be restrained to avoid diffusing out of the binding site and ensuring phase-space overlap so that intermediate configurations smoothly connect ligand basins. However the number and definition of restraints should be chosen carefully as they can introduce bias in the simultations if not explicitely corrected. Multistate hybrid topologies can alleviate this issue as well as improve phase-space overlap \cite{champion2023accelerating} but their implementation in most MD engines usually requires custom routines, limiting their applicability.

RE-EDS workflows as implemented in the GROMOS \cite{champion2023accelerating} or OpenMM \cite{rieder2022leveraging} software packages include an automated calibration procedure in which the EDS potentials on the replica-exchange ladder are iteratively optimized by minimizing round-trip times and state occupancy. Nevertheless, heterogeneous datasets containing ligands that necessitate many heavy atom insertions or deletions can drastically reduce the efficiency of RE-EDS, and in these cases it is generally beneficial to split very large or poorly overlapping ligand sets into smaller groups or subsets to be executed with pairwise methods.
}

\paragraph{Perturbation maps}
\tocsectioncomment{Perturbation maps paragraph: @Mary Pitman/ @Jenke Scheen/ OpenFE team to update discussion of perturbation maps?}
Based on the input ligand series, a perturbation map or network can be planned. Recent heuristics have shown the more connected the perturbation network the better. However, there is a way to optimize network structure while minimizing the number of perturbations that need to be computed reducing the resulting computational cost~\cite{yang2020optimal,xu2019optimal}. Sometimes the introduction of intermediates that are not part of the original congeneric series are essential to avoid ring breaking, or to deal with perturbations that would otherwise result in large numbers of atoms being inserted or deleted. Some commercial tools have good underlying heuristics but may fail with complicated input, needing user validation in particular when dealing with chiral compounds.

@@

In some cases, during the lead optimization stage, or for very large datasets that would benefit from rougher initial free energy ranking, or in cases where perturbations would be rather large, a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} \textbf{A} is used. However, adding redundancy into the network means that a better error analysis can be carried out by looking at cycle closure errors as discussed in sec.~\ref{sec:are-they-good}, with an example given in Fig.~\ref{fig:fig_types_of_networks}\textbf{B}.

Methods from experimental design have been applied to the construction of the perturbation maps. Yang et al.~\cite{yang2020optimal} showed how to optimize the perturbation map by selecting a fixed number of calculations from the pairwise perturbations so that the resulting set of calculations minimize the total variance. Xu~\cite{xu2019optimal} showed how to optimize the perturbation map by allocating different amounts of simulation time to different pairwise perturbations so as to minimize the total variance, given the total simulation time of all the perturbation calculations. Both approaches lead to substantial reduction in the statistical error of the estimated free energies.
